---
title: exploring NUFORC sightings
author: Andrew
date: '2017-07-08'
slug: exploring-nuforc-sightings
categories:
tags: ["R", "data analysis", "data viz", "UFOs", "kaggle"]
---

```{r opts, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.align = "center")
```

```{r init}
library(tidyverse)
library(scales)
library(lubridate)
library(magrittr)
library(plotly)
library(broom)
library(leaflet)
library(leaflet.extras)
library(tidycensus)
library(sf)

options(tigris_use_cache = TRUE)

# we'll look at the 'scrubbed' data, no incomplete reports for now
scrubbed <- read_csv("../../static/data/ufo-sightings/scrubbed.csv.zip")
```

```{r prep}
# let's look only at US sightings
us <- scrubbed %>% 
  filter(country == "us") %>%
  mutate(
    date_time    = mdy_hm(datetime),
    year_sighted = year(date_time),
    date_posted  = as.Date(`date posted`, "%m/%d/%Y"),
    duration     = `duration (seconds)`
  )

us <- filter(us, year_sighted < 2014)
```

*R code used for each of the graphics is available [here.](https://github.com/mooreaw/ufo-sightings)*

<!-- https://www.kaggle.com/NUFORC/ufo-sightings -->

While flipping through an issue of *the Economist* a few years ago, I stumbled across an [article](http://www.economist.com/news/united-states/21605918-everything-you-need-know-about-ufos-0) summarizing UFO sightings reported across the US. It wasn't a full feature, but the topic was playful and I lingered on it longer than I spent with the rest of the issue. The author had put together a map charting sightings by state, and a plot looking at how sightings were distributed across the day. Their conclusions were pretty straightforward:

1. sightings are more likely to be rural, "shunning" big cities due to light pollution

2. sightings are most likely to occur
    - when people aren't working 
    - when people are likely to be drinking

For a short blurb, these takeaways are tidy, but for a piece titled "Everything you need to know about UFOs", I wanted more to chew on. How much is alcohol really implicated in sightings of this kind? What about other local factors? These felt like glaring questions, but my musing was shelved when I closed the magazine. Luckily, Kaggle posted an updated version of the dataset to [download](https://www.kaggle.com/NUFORC/ufo-sightings), and after finding it, I couldn't wait to look at some of these questions closer.

Each row in the data is an individually reported sighting, archived by the National UFO Research Center in Washingon [(NUFORC)](http://www.nuforc.org/). Over 80,000 events are recorded, going all the way back to the early 1900s. In addition to listing the date/time that a sighting was reported, the shape of the object, city, state, country, and latitude/longitude coordinates are all recorded. Most of the reports are from the US (`r percent(nrow(us) / nrow(scrubbed))`), so I decided to keep my focus there.

### the number of sightings reported is increasing

As far as I know, national excitement about UFOs was strongest during the Cold war. [Project Blue Book](https://en.wikipedia.org/wiki/Project_Blue_Book) was set up by the US Airforce in the 50s, and aimed to collect evidence and explain different sightings being reported. It was ultimately shut down due to inconclusive findings. I expected to see more documented sightings between the 50s-70s, but this definitely isn't reflected in the file. This is a good point to consider the limitations of what we're working with. NUFORC was started in 1974, so perhaps it's surprising they've taken the time to record anything earlier than that at all. Someone would have to be pretty motivated to call-in about an experience they had years after it happened. Second, even closer to the present, it seems safe to say that NUFORC's data doesn't capture all the events that might be called sightings on an annual basis.

That said, the number of sightings in each state has seen a steady climb since 2000. My guess is that due to prominence of search engines, awareness of NUFORC (and thus, the likelihood of reporting) has increased. You can follow the trend with the boxplots below. Each state contributes their count for a given year, and the boxplot helps give a sense how much they vary from each other.

```{r medians-over-time}
by_year_boxplot <- us %>% 
  count(year_sighted, state) %>% 
  ungroup() %>% 
  filter(year_sighted > 1990) %>% 
  ggplot(aes(x = factor(year_sighted), y = n)) + 
  geom_boxplot(fill = "steelblue") +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 90)
  ) +
  labs(x = "", y = "# of Sightings", title = "Median Number of State Sightings, 1991-2013")

ggplotly(by_year_boxplot)
```
*(The outliers from each year reliably include a handful of states from the west, like Washington and California.)*

### what people report seeing

One of the things I was most curious about that went unmentioned in the *Economist* piece was what people say they're seeing! The next chart looks at the composition of all US sightings reported during each year. The thicker the color band, the higher the percentage of a year's sightings can be attributed to a set of shapes. Across the whole time span, the most common kind of UFO are lights (maybe looking something like [this?](https://youtu.be/qxLEnqsUYS4)), but this wasn't always the case. Earlier sightings were much more likely to report the classic saucer/disk shape. The fact that we don't have as many reports from the 60s and 70s is a little disappointing. Because we're simply seeing thousands more sightings in recent years, it's hard to say why the disk shape has fallen out of favor. Air travel has also become increasingly ubiquitous since the 70s, so perhaps the increasing share of "lights" sightings can be attributed to that.

```{r shapes, fig.height=8.5, fig.width=10.5}
shape_lut <- c(
  "cylinder"  = "cylinders, cigars, crescents, cigars, crosses",
  "circle"    = "circles, spheres, ovals, teardrops, eggs, round",
  "light"     = "lights, fireballs, flashes, formations",
  "sphere"    = "circles, spheres, ovals, teardrops, eggs, round",
  "disk"      = "disk",
  "fireball"  = "lights, fireballs, flashes, formations",
  "unknown"   = "unknown, other, changing",
  "oval"      = "circles, spheres, ovals, teardrops, eggs, round",
  "other"     = "unknown, other, changing",
  "rectangle" = "rectangle, hexagon",
  "chevron"   = "chevrons, deltas, triangles, pyramids, diamonds, cones",
  "formation" = "lights, fireballs, flashes, formations",
  "triangle"  = "chevrons, deltas, triangles, pyramids, diamonds, cones",
  "cigar"     = "cylinders, cigars, crescents, cigars, crosses",
  "NA"        = "missing",
  "delta"     = "chevrons, deltas, triangles, pyramids, diamonds, cones",
  "changing"  = "unknown, other, changing",
  "diamond"   = "chevrons, deltas, triangles, pyramids, diamonds, cones",
  "flash"     = "lights, fireballs, flashes, formations",
  "egg"       = "circles, spheres, ovals, teardrops, eggs, round",
  "teardrop"  = "circles, spheres, ovals, teardrops, eggs, round",
  "cone"      = "chevrons, deltas, triangles, pyramids, diamonds, cones",
  "cross"     = "cylinders, cigars, crescents, cigars, crosses",
  "pyramid"   = "chevrons, deltas, triangles, pyramids, diamonds, cones",
  "round"     = "circles, spheres, ovals, teardrops, eggs, round",
  "flare"     = "lights, fireballs, flashes, formations",
  "hexagon"   = "rectangle, hexagon",
  "crescent"  = "cylinders, cigars, crescents, cigars, crosses",
  "changed"   = "unknown, other, changing"
)

us$shape_simp <- shape_lut[us$shape]

shapes_frame <- us %>% 
  mutate(shape_simp = fct_explicit_na(shape_simp, "missing")) %>% 
  count(year_sighted, shape_simp) %>%       # need to fill years without all the
  spread(shape_simp, n, fill = 0) %>%       # levels with 0s to make sure the plot
  gather(shape_simp, n, -year_sighted) %>%  # doesn't get messed up
  group_by(year_sighted) %>%
  mutate(pct  = n / sum(n)) %>%
  ungroup() %>%
  arrange(year_sighted, pct) %>%
  filter(year_sighted > 1949)

shapes_frame$shape_simp <- shapes_frame %$%
  fct_reorder(factor(shape_simp), n, .desc = TRUE)

shapes <- shapes_frame %>% 
  filter(year_sighted != 1995, year_sighted != 1996) %>% 
  ggplot(aes(x = year_sighted, y = pct, fill = shape_simp, group = shape_simp)) +
  geom_area(aes(fill = shape_simp, group = shape_simp), alpha = .8) +
  theme_minimal(base_size = 15) +
  scale_y_continuous(label = percent) +
  guides(fill = guide_legend(nrow = 4)) +
  theme(
    panel.grid.minor = element_blank(),
    legend.title     = element_blank(),
    legend.position  = "bottom"
  ) +
  labs(
    x = "", y = "", 
    title = "USA: UFO Shapes by Year, 1950 - 2013", 
    caption = "* 1995 & 1996 excluded due to high levels of missing data."
  ) +
  scale_fill_brewer(palette = 16)

shapes
```

### seasonal differences

Next, I wanted to know whether there's a particular "season" for spotting UFOs. To get there, I created a rate of sightings per million people in the US population. This is a point where we can get a grip on just how rare sightings are, per-person-- even in 2013 we're only seeing 1-2 sightings each month per million people. You can see sightings become more common as we move toward the present, but there are also defined portions of the year that people are most likely to report sightings-- namely during the summer and fall. This lines up with what the *Economist* reported (i.e. sightings happen most frequently during evenings and weekends), given that these parts of the year tend to be when people spend the most time outside.

```{r usa-yr-month, fig.height=5.5, fig.width=10}
# dataset from tidyr
data(population)

uspop <- filter(population, country == "United States of America")

p2 <- us %>% 
  count(mo = month(date_time), year_sighted) %>% 
  inner_join(uspop, by = c("year_sighted" = "year")) %>% 
  mutate(
    Rate  = n %>% divide_by(population) %>% multiply_by(1000000),
    Month = factor(mo, levels = 1:12, labels = month.abb),
    Year  = factor(year_sighted, labels = c("1995", str_c("'", c(96:99, str_c("0", 0:9), 10:13))))
  ) %>% 
  ggplot(aes(x = Year, y = Month, fill = Rate)) +
  geom_tile() +
  theme_minimal(base_size = 15) +
  scale_fill_gradient2(low = "white", high = "blue") +
  theme(legend.title = element_blank()) +
  labs(x = "", y = "", title = "UFO Sightings in the USA per 1,000,000 People")

ggplotly(p2)
```

### where are sightings most likely to happen?

As I mentioned above, the *Economist* piece suggests that sightings tend to spurn big cities, rationale being that light pollution drives down reports. They also suggested that sightings favor the US/Canada border, and other rural areas. I'm not sure if I've reached the same conclusions... The map below reflects two pieces of data for each county in the US during 2013: population density (people/km^2), and sightings reported. The little grid below reflects how I've categorized each county. The top-left (dark-grey) represents counties with a really high population density but low numbers of sightings. The bottom right (deep brown) represents counties with a low population density but really high sighting count. Up in the top right we have counties whose number of sightings tracks well with their population size. 
 
```{r build-bv-map, cache = TRUE}
us$geometry <- 1:nrow(us) %>%
  lapply(function(x) st_point(c(us$longitude[x], us$latitude[x]))) %>% 
  st_sfc(crs = 4269)

us2 <- st_sf(us)

countydat <- 
  get_acs(
    geography = "county",
    variables = c("B17006_001", "B17006_002"),
    output    = "wide",
    geometry  = TRUE
  ) %>%
  mutate(area = st_area(geometry)) %>% 
  st_transform(4269) %>% 
  st_join(us2, join = st_contains)

rates <- countydat %>%
  as_tibble() %>%
  count(NAME) %>% 
  right_join(
    countydat %>% 
      select(NAME, B17006_001E, area, geometry) %>%
      as_tibble() %>%
      distinct(NAME, .keep_all = TRUE)
  ) %>%
  mutate(
    d = B17006_001E / as.numeric(area / 1000000),
    dm = Hmisc::cut2(d, g = 3, levels.mean = TRUE),
    nm = Hmisc::cut2(n, g = 3, levels.mean = TRUE)
  )

levels(rates$dm) <- 1:3
levels(rates$nm) <- 1:3

rates$bin <- str_c(rates$dm, "-", rates$nm) %>% 
  factor(levels = c(
    "3-1", "2-1", "1-1",  # col 1 -->
    "3-2", "2-2", "1-2",  # col 2 -->
    "3-3", "2-3", "1-3"   # col 3 -->
  ))

vals <- c(
  "#8c510a", "#bf812d", "#dfc27d", # col 1 -->
  "#f6e8c3", "#f5f5f5", "#c7eae5", # col 2 -->
  "#80cdc1", "#35978f", "#01665e"  # col 3 -->
)

pal <- colorFactor(palette = rev(vals), rates$bin)

county_poly <- rates %>%
  filter(!str_detect(NAME, "Alaska|Hawaii|Puerto")) %>%
  st_sf() %>% 
  st_transform(crs = "+proj=longlat +datum=WGS84")

county_ctr <- county_poly %>%
  st_cast("POLYGON") %>%
  st_centroid()

bv_choro <- leaflet() %>%
  addPolygons(
    data = county_poly,
    color = "black",
    popup = ~str_glue_data(., "{NAME}; Sightings: {n}, Density: {round(d, 2)}"),
    stroke = TRUE,
    weight = 1,
    smoothFactor = 0.2,
    fillOpacity = 1,
    fillColor = ~ pal(bin),
    highlightOptions = highlightOptions(color = "white", weight = 2, bringToFront = TRUE)
  ) %>%
  addResetMapButton()

legend_p <- rates %>% 
  select(dm, nm, bin) %>% 
  distinct() %>% 
  arrange(bin) %>% 
  ggplot(aes(x = nm, y = dm, fill = bin)) +
  geom_tile() +
  scale_fill_manual(values = rev(vals)) +
  theme_minimal() +
  labs(x = sprintf("Sightings \u2192"), y = sprintf("Density \u2192")) +
  theme(
    panel.grid = element_blank(),
    axis.text  = element_blank(),
    legend.position = "none"
  )

# rates %>% 
#   group_by(bin) %>% 
#   arrange(desc(n), desc(d)) %>% 
#   slice(1) %>%  
#   select(bin, n, d) %>% 
#   arrange(bin)
```
<center>
```{r print-map}
bv_choro
```
</center>

```{r print-legend, fig.height=1.5, fig.width=1.5}
legend_p
```

The thing I find striking is that there are a few really urban counties that show a large number of sightings. Cook (Chicago, IL), Wayne (Detroit, MI), King (Seattle, WA), New York, and San Francisco all follow this pattern. We do see a fair amount of rural counties with disproportionate levels of sightings in the west and northeast, but I'm not fully sold on the "shunning" asserted in the *Economist* piece. Additionally, the central US (including the borders) appears fairly neglected as far as sightings go. This makes me feel the urban/ruralilty relationship isn't super straightforward.

### Are there cultural or population-based factors that are related to sightings?

```{r sightings-glm-0}
census <- get_acs(
  geography = "state", 
  endyear = 2013,
  variables = c(
    "B25064_001",  # median gross rent
    "B06011_001",  # median income
    "B01002_001",  # median age
    "B02001_003",  # black alone
    "B02001_002",  # white alone
    "B03002_012",  # hispanic or latino
    "B02001_005",  # asian alone
    "B16010_028",  # some college or associates
    "B16010_041"   # bachelor's or higher
  ),
  summary_var = "B01003_001"      # total people in unit
)

vlut <- c(
  "B25064_001" = "median gross rent",
  "B06011_001" = "median income",
  "B01002_001" = "median age",
  "B02001_003" = "black alone",
  "B02001_002" = "white alone",
  "B03002_012" = "hispanic or latino",
  "B02001_005" = "asian alone",
  "B16010_028" = "some college or associates",
  "B16010_041" = "bachelors or higher"
)

census$variable <- vlut[census$variable]

medians <- census %>%
  filter(str_detect(variable, "median")) %>%
  select(NAME, variable, estimate, total_population = summary_est) %>%
  mutate(variable = str_replace_all(variable, " ", "_")) %>%
  spread(variable, estimate)

props <- census %>%
  mutate(variable = str_replace_all(variable, " ", "_")) %>%
  filter(!variable %in% names(medians)) %>%
  mutate(prop = estimate %>% divide_by(summary_est) %>% multiply_by(100)) %>%
  select(NAME, variable, prop) %>%
  spread(variable, prop)
```

```{r st-binge, eval = FALSE}
library(rvest)

ele <- "table:nth-child(4) td , table:nth-child(4) th"

binge <- "https://www.cdc.gov/alcohol/data-table-text.htm#prevalence" %>% 
  read_html %>% 
  html_nodes(ele) %>% 
  html_text()

binge_clean <- tibble(txt = binge) %>% 
  filter(!str_detect(txt, "\\("), txt != "") %>% 
  mutate(txt = str_c(txt, "-", lead(txt))) %>% 
  filter(!str_detect(txt, "^[[:digit:]]"), !str_detect(txt, "%")) %>% 
  separate(txt, c("state", "pct"), "-") %>% 
  mutate(pct = as.numeric(pct))

write_csv(binge_clean, "../../static/data/ufo-sightings/2015-binge-prev-cdc.csv")
```

```{r sightings-glm-1}
demog <- medians %>%
  left_join(props, by = "NAME") %>% 
  rename(state = NAME)

binge <- read_csv("../../static/data/ufo-sightings/2015-binge-prev-cdc.csv")
b5tsc <- read_csv("../../static/data/ufo-sightings/rentfrow-state-t-scores-b5.csv")
relig <- read_csv("../../static/data/ufo-sightings/wiki-relig.csv")

relig$region <- str_to_title(relig$region)

binge <- rename(binge, binge_prev = pct)

yr13 <- us %>% 
  filter(year_sighted == 2013) %>% 
  count(state.abb = str_to_upper(state)) %>%
  rename(sightings = n) %>% 
  left_join(tibble(state.abb, state = state.name), "state.abb")

demog <- medians %>%
  rename(state = NAME) %>%
  left_join(props, c("state" = "NAME")) %>%
  left_join(b5tsc, "state") %>% 
  left_join(binge, "state") %>% 
  left_join(relig, c("state" = "region")) %>% 
  left_join(yr13, "state") %>% 
  filter(!is.na(sightings))

sight_desc <- summarise(demog, avg = mean(sightings), sd = sd(sightings))

demog <- demog %>%
  rename(religiosity = pct) %>%
  mutate(
    pop_norm_z = (total_population - mean(total_population)) / sd(total_population),
    rent_norm_z = (median_gross_rent - mean(median_gross_rent)) / sd(median_gross_rent),
    income_norm_z = (median_income - mean(median_income)) / sd(median_income)
  )
```

What about other local characteristics? The author in the *Economist* piece seemed to suggest that drunk (and probably rural) people are the ones stumbling around and reporting these things. I'm not inclined think alcohol consumption wouldn't be related to a person experiencing a "sighting", but maybe we can try to bring data of some kind to bear on the question. At the same time, states and regions aren't (solely) reducible to their drinking habits, so I looked at some other candidate variables. In order to examine these relationships, I built a regression model with state-based counts of sightings from 2013. I decided to fit a Poisson regression, given that our dependent variable is a count (visually the distribution seems to match, but the variance is bigger than the mean, indicating some overdispersion). With our dependent variable set, I looked at the following group of predictors:

1. *US Census Estimates:* 
    - population size, median income, and median cost of rent (converted to Z-scores)
    - percent of total population recorded as white, black, Asian, and Hispanic/Latino
    - percent of total population with a bachelor's degree or higher
    - median age
2. *Religiosity:*
    - gathered from [Wikipedia](https://en.wikipedia.org/wiki/List_of_U.S._states_by_religiosity), reported by Gallup polling
    - percentage of sampled individuals who rate religion as 'important' or 'very important'
3. *CDC prevalence of binge-drinking:*
    - data available [online](https://www.cdc.gov/alcohol/data-stats.htm)
    - state prevalence (%) of binge/excessive drinking estimated using data from the Behavioral Risk Factor Surveillance System (BRFSS) survey
4. *State Big-5 Personality Traits from Rentfrow et al. (2013):*
    - publication available [online](https://www.apa.org/pubs/journals/releases/psp-a0034434.pdf) from the American Psychological Association
    - T-Scores for each state on the [Big-5 personality traits](https://en.wikipedia.org/wiki/Big_Five_personality_traits), *openness, conscientiousness, extraversion, agreeableness, and neuroticism* based on aggregated online samples from over 1.5 million participants

While the census data is current for 2013, the other variables were produced/collected between 2013-2015. Not ideal, but I'm assuming their values are probably pretty stable from year to year. Another consideration is that the model might perform better if we could look at a more detailed observation level (e.g. counties or census tracts within states), but aside from the census data, the other characteristics are a bit harder to come by. Anyways, let's take a quick look to see if any glaring characteristics stand out with the variables we have:

```{r var-facet}
levs <- c(
  "pop_norm_z", "income_norm_z", "rent_norm_z", "median_age", "bachelors_or_higher", 
  "black_alone", "white_alone", "hispanic_or_latino", "asian_alone",
  "binge_prev", "religiosity",
  "o", "c", "e", "a", "n"
)

labs <- c(
  "Population (Z-Score)", "Median Income (Z-Score)", "Median Rent (Z-Score)",
  "Median Age", "% Bachelors or higher",
  "% pop. Black-Alone", "% pop. White-Alone", "% pop. Hispanic/Latino", "% pop. Asian",
  "Binge Drinking", "% Religious Important",
  "Openness", "Conscientiousness", "Extraversion", "Agreeableness", "Neuroticism"
)

scatter_dat <- demog %>% 
  select(
    state, 
    sightings,
    contains("_alone"),
    contains("or"), e:o, binge_prev, religiosity, median_age,
    -contains("some_college")
  ) %>% 
  gather(var, val, -state, -sightings) %>% 
  mutate(var = factor(var, levs, labs))

scatter_facet <- scatter_dat %>% 
  ggplot(aes(x = val, y = sightings)) +
  geom_point() +
  facet_wrap(~var, scales = "free") + 
  geom_smooth(method = "glm", method.args = list(family = "poisson")) +
  theme_minimal() +
  theme(panel.grid.minor = element_blank()) +
  labs(x = "", y = "# Sightings")

scatter_facet
```

In each plot, every state is represented as a single dot, with the number of sightings increasing along the y-axis, and the variable of interest along the x-axis. I've added a smoother to each plot to help summarize the relationships. Flat lines suggest small or negligible associations. This seems to be largely the case for most of our characteristics, but, population size and cost of rent might be showing a hint of something. Of note, the binge-drinking prevalence variable doesn't appear to be related to the number of sightings at all. We can be a little more specific by letting the regression model simplify some of this for us.

```{r}
# performs a poisson regression, modeling sightings as a function of 
# state characteristics
# http://stats.stackexchange.com/questions/18480/interpretation-of-log-transformed-predictor
sightings_pois <- glm(
  sightings ~
    pop_norm_z  +
    white_alone +
    black_alone +
    asian_alone +
    hispanic_or_latino +
    rent_norm_z +
    median_age +
    income_norm_z +
    bachelors_or_higher +
    religiosity +
    binge_prev +
    o +
    c +
    e +
    a +
    n
  ,
  data = demog,
  family = poisson
)

roundy <- function(x) round(x, 2)

# http://www.biostat.umn.edu/~dipankar/bmtry711.11/lecture_13.pdf
# http://stats.idre.ucla.edu/stata/output/poisson-regression/
# exponentiating the betas gives you the *incident rate ratio*
# poisson regression coefficients are the differences between
# the log of expected counts, i.e.:
# log(y x+1) - log(y x) or -> log(y x+1 / y x)
#           ^ difference of two logs is equal to their quotient
# IRR: the outcome (y) is multiplied by the beta
# i.e., in the case of our model, if openness increases by 1 point,
# the number of sightings increases by 1.03.
sightings_est <- sightings_pois %>%
  tidy() %>%
  bind_cols(confint_tidy(sightings_pois)) %>%
  select(term, estimate, contains("conf")) %>%
  mutate_each(funs(roundy(exp(.))), -term) %>%
  arrange(desc(abs(1 - estimate)))
```

```{r fitted-values-plot}
p_fitted <- augment(sightings_pois, type.predict = "response") %>%
  ggplot(aes(x = .fitted, y = sightings)) +
  geom_abline(slope = 1, intercept = 0, lty = "dashed") +
  geom_point() +
  theme_minimal(base_size = 15) +
  theme(panel.grid.minor = element_blank()) +
  labs(
    x = "Fitted (Predicted) Value",
    y = "# of Sightings (2013)"
  )

p_fitted

pois_rmse <- yardstick::rmse(p_fitted$data, sightings, .fitted)
pois_mae <- yardstick::mae(p_fitted$data, sightings, .fitted)
```

So, how well did the model do? Not perfect, but also not entirely terrible? Above is a fitted-value plot; you can see that the model has overestimated sighting counts to varying degrees for about half the states. Just under half the states have sighting counts under 100 for 2013, and it looks like a lot of our misses were in that region. The RMSE and MAE for this model are `r round(pois_rmse$.estimate, 1)` and `r round(pois_mae$.estimate, 1)`, respectively.

```{r model-results}
tidy(sightings_pois, conf.int = TRUE) %>%
  mutate(
    across(.cols = matches("estimate|conf"), .fns = exp),
    term = factor(term, levs, labs),
    term = fct_reorder(term, 0 - estimate),
    term = fct_rev(term)
  ) %>%
  filter(term != "(Intercept)") %>%
  ggplot(aes(x = term, y = estimate, ymin = conf.low, ymax = conf.high)) +
  geom_pointrange() +
  geom_text(aes(label = ifelse(estimate > 1.1, as.character(round(estimate, 2)), "")), vjust = 1.6) +
  geom_text(aes(label = ifelse(estimate < 0.8, as.character(round(estimate, 2)), "")), vjust = -1.3) +
  coord_flip() +
  theme_minimal(base_size = 15) +
  theme(panel.grid.major.y = element_blank()) +
  labs(x = "", y = "IRR (& 95% CIs)", title = "Estimated IRRs for each predictor")
```

Okay, last plot, I promise! A major benefit of a regression model is that we can see how individual variables contribute to the outcome we're interested in. The figure above shows *Incident Rate Ratios* (IRR) for each of the predictors included in the model. This estimate shows the *multiplicative* change in the dependent variable, for each 1-unit increase in a predictor. If we imagine we could pluck out an "average" state, and magically increase its median age by 1 year, we'd expect its number of sightings to increase by 10%. What we're seeing is that income, rent, and population size appear to be related with our outcome. Everything else looks pretty negligible.

Wrapping up, we have to use a lot of caution given our unit of analysis. Sightings are experienced by persons, not states. I wanted to see if a state-level summary of alcohol (over) consumption was related to a state's recorded count of sightings, and even though the estimated association is basically nil, this doesn't rule out the notion that intoxicated people are more likely to experience sightings. It could be that the population of people logging onto NUFORC's website to report sightings tend to drink more, but we don't have any information about this group at-hand. That said, I think that we can see from both the choropleth and this higher level of aggregation that population-size seems to matter. There seem to be regional differences, and rural areas don't appear to be created equally in terms of the frequency of sightings. In any case, I hope you enjoyed this blog post! This dataset is quirky and fun to explore. It might make a good addition to any future workshops on data visualization. `r emo::ji("smile")`
