---
title: guessing how many streak-freezes I'll use over the next 4 months
author: Andrew
date: '2021-05-01'
slug: estimating-streak-freezes-duolingo
categories: []
tags: [language learning, statistics, simulation]
---

```{r opts, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, fig.align = "center")
```

So, this is a small thing, but something I'm proud of. Over the past year, I've been practicing German (`r emo::ji("de")` *Ich habe Deutsch gelernt!*) using Duolingo's mobile app. I don't have an especially romantic reason for why I settled on the language, but I had watched the first season of *Dark* on Netflix in 2018, and don't really enjoy dubbed foreign film/tv. Listening to the language prompted me to try a few lessons initially, but I settled into routine practice while finishing *Babylon Berlin*[^1] last fall. I don't remember setting an explicit goal to reach a year of daily practice, but we're coming up on that point. According to the app, I've been on a streak of 217 days! *Leider habe ich nicht jeden Tag geübt.* There were several days where I missed my usual 20-25 minutes of practice, but Duolingo lets you purchase "streak freezes" with in-app currency to preserve your progress. Here's a snapshot of my streak(s) over the past 8 months.

```{r start}
library(tidyverse)
library(lubridate)
library(knitr)

set.seed(20210430)

theme_set(theme_minimal(base_size = 15) + theme(panel.grid.minor = element_blank()))

dates <- tibble(date = as.Date(mdy("8/31/20"):mdy("4/30/21"), origin = "1970-1-1")) %>%
  left_join(
    tibble(
      date   = mdy(c("10/1/20", "10/4/20", "10/15/20", "10/18/20", "10/20/20", "10/21/20", "10/25/20", "11/3/20", "11/25/20", "12/23/20", "12/25/20", "1/1/21", "1/2/21", "1/12/21", "1/17/21", "1/19/21", "1/27/21", "2/7/21", "2/16/21", "2/20/21", "2/23/21", "2/25/21", "3/4/21", "3/12/21", "3/14/21", "4/14/21")),
      freeze = 1L
    ),
    by = "date"
  ) %>%
  mutate(
    freeze = replace_na(freeze, 0L),
    mon = factor(
      month(date),
      levels = c(8:12, 1:4),
      labels = c(str_c(month.abb[8:12], " '20"), str_c(month.abb[1:4], " '21"))
    ),
    wk = case_when(
      date == first(date) ~ 1L,
      date == last(date)  ~ 32L,
      TRUE ~ cumsum(lead(week(date)) != week(date))
    )
  )
```

```{r step plot}
ggplot(dates, aes(x = date, y = cumsum(freeze))) +
  geom_step(color = scico::scico(1, palette = "nuuk")) +
  geom_point(
   data = tibble(date = mdy(c("10/20/20", "12/23/20", "1/17/21", "2/20/21", "3/14/21")), freeze = seq(5, 25, 5)),
   aes(x = date, y = freeze), color = scico::scico(1, palette = "nuuk"), size = 2
  ) +
  scale_x_date(date_breaks = "months", date_labels = "%m/%y", expand = c(0.1, 0)) +
  labs(x = "", y = "Cumulative Freezes Used") +
  theme_minimal(base_size = 15) +
  theme(panel.grid.minor.x = element_blank()) +
  annotate("text", x = mdy("9/4/20"), y = 2, label = 'Watching\n"Babylon Berlin"') +
  annotate("text", x = mdy("12/6/20"), y = 4, label = "Extra motivated?\nkeine Ahnung") +
  annotate("text", x = mdy("1/1/21"), y = 20, label = "Distracted, 11 freezes\naccumulated during\nJan. & Feb.") +
  annotate("text", x = mdy("4/20/21"), y = 23, label = 'Finishing "Dark"\non Netflix')
```

So, each freeze is a day that I'm "behind" on my goal. If I've used `r sum(dates$freeze)` freezes now, how many more should I expect over the next 4-ish months? I started right around the end of August, so 17 weeks from now would be close to my 1-year mark of daily practice. When counting the number of freezes on a weekly basis, the distribution looks fairly close to a Poisson distribution.

```{r distribution, echo = TRUE}
# I've omitted my prep code-- 'dates' is a dataset with 1 row/day,
# and an indicator 0/1 for whether a streak-freeze was used on a given day.
data.frame(dates)[1:3, ]

weekly <- dates %>%
  group_by(wk) %>%
  tally(freeze)

weekly %>%
  summarise(wk = max(wk), m = mean(n), v = var(n), min = min(n), max = max(n)) %>%
  kable(col.names = c("# Weeks", "Mean", "Variance", "Min.", "Max."), digits = 3)

library(distributions3)

compare_w_poisson <- weekly %>%
  count(n) %>%
  mutate(p = nn / sum(nn), poi = pmf(Poisson(0.722), n)) %>%
  pivot_longer(p:poi)

ggplot(compare_w_poisson, aes(x = n, y = value, fill = name)) +
  geom_col(position = "dodge") +
  geom_text(aes(label = round(value, 2)), position = position_dodge(1), vjust = -0.25) +
  scico::scale_fill_scico_d(name = "", palette = "nuuk", labels = c("p" = "Observed Proportion", "poi" = "Poisson Probability (&lambda; = 0.722)")) +
  labs(x = "# freezes used during a given week", y = "") +
  theme(legend.position = "top", legend.text = ggtext::element_markdown())
```

Maybe my data isn't a perfect fit to a Poisson distribution with the same mean, but perhaps it's close enough to serve as a model for what we can expect. So, to be specific, let's let

<center>
$X$ = the number of streak freezes used in a 7-day period, where $X \sim Poisson(\lambda = 0.722)$.
</center>
<br>

We can then simulate 17 weeks from $X$ and sum the results, repeating this process say, 10,000 times. Or, more formally, we end up with a vector of sums $\vec{s}$: 

<center>
$\vec{s} = \begin{bmatrix}s_1 \\ s_2 \\ \vdots \\ s_j \end{bmatrix} \text{ where } j = 1, 2, 3, \cdots, 10,000, \text{ and } s_j = \sum_{i=1}^{17}[\vec{x}_i]$, if we let $\vec{x}$ be the $j^{th}$ sample from $X$.
</center>
<br>

For this simulation, we're assuming that the results of each week are independent of each other. This feels reasonable to me; autocorrelation in my tabulated weekly counts seems negligible.[^2] All that's left is to set up a loop to collect the simulation results, and then we'll use a histogram to visualize them. <!-- acf(weekly$n, plot = FALSE) --> 

```{r sim, echo = TRUE}
X <- Poisson(0.722)
s <- c()

for (b in 1:10000) {
  x <- random(X, n = 17)
  s <- c(s, sum(x))
}

ggplot(tibble(s), aes(x = s)) +
  geom_histogram(color = "white", bins = 15) +
  scale_y_continuous(labels = scales::comma) +
  theme(plot.caption = ggtext::element_markdown()) +
  labs(
    x = "Total # of simulated freezes observed over 17 weeks", y = "N",
    caption = str_glue(
      "*s*<sub>Mean</sub> = {round(mean(s), 2)}, *s*<sub>SD</sub> = {round(sd(s), 2)} ",
      "Pr(8 &le; *s* &le; 13) = {sum(between(s, 8, 13)) / length(s)}<br>",
      "Pr(7 &le; *s* &le; 21) = {sum(between(s, 7, 21)) / length(s)}"
    )
  )
```

We end up with a fairly normal-looking histogram, as would be expected by the central limit theorem.p[^3] If the model is appropriate, it seems like I should expect between 8 to 13 additional freezes to be accumulated over this time period. The simulation results suggest there's only a `r scales::percent(sum(s < 10) / length(s))` chance that the number of freezes accumulated will be less than 10. Pulling everything together, by the end of August I'll probably be between 34 and 39 streak-freezes deep. This means it'll be at least a month after my starting point before I can truly claim I've met my goal. `r emo::ji("sob")`

[^1]: Which I recommend if you're into *noir*, but the tragic & foreshadowed nature of the historical setting is captivating on its own. The soundtrack for each season has been excellent as well.

[^2]: Using a lag of up to 15 weeks, the autocorrelations (assessed by `acf()`) ranged between 0.15 to -0.25, but most were much smaller in terms of their absolute magnitude.

[^3]: update/edit: in the process of wrapping up this post, I came across this [question/answer](https://math.stackexchange.com/a/451996) on SO, which suggests that my distribution here is actually *Poisson*, not normal. In theory we're actually looking at a new Poisson distribution with $\lambda = 0.722 \times 17 = 12.274$, which lines up close to the sample mean from the simulation (the sample variance is a little off, but this is probably to be expected from the randomness of the simulation). I guess I can save myself a step in the future, but working through things was a good learning experience.
